Ethnicity sensitive author disambiguation using semi-supervised learning

### 2016-01-11 paper/author_disambiguaton.tex
correct i.e. and e.g.

@@

\section{Introduction}
In academic digital libraries, author name disambiguation is the problem of
grouping together publications written by the same person.
Author name disambiguation is often a difficult problem because an author may use different
spellings or name variants across their career (synonymy) and/or distinct authors may
share the same name (polysemy).
Most notably, author disambiguation is often more troublesome for researchers
from non-Western cultures, where personal names may be traditionally less diverse (leading to
homonym issues) or for which transliteration to Latin characters may not be unique (leading to
synonym issues).
With the fast growth of the scientific literature, author disambiguation has become a pressing
issue since the accuracy of information managed at the level of individuals directly affects:
the relevance search of results (e.g., when querying for all publications written by a given author);
the reliability of bibliometrics and author rankings (e.g., citation counts or other impact
metrics, as studied in \cite{strotmann2012author}); and/or the relevance of scientific network analysis
\cite{newman2001structure}.

Efforts and solutions to author disambiguation have been proposed from various
communities \cite{liu2014author}. On the one hand, libraries have maintained
authorship control through manual curation, either in a centralized way by
hiring professional collaborators or through developing services that invite authors to register
their publications themselves (e.g., Google Scholar or Inspire-HEP).
Recent efforts to create persistent digital identifiers assigned to researchers (e.g., ORCID or ResearcherID),
with the objective to embed these identifiers in the submission workflow of publishers
or repositories (e.g., Elsevier, arXiv or Inspire-HEP), would univocally solve any disambiguation issue.
With the large cost of centralized manual authorship control, or until crowdsourced solutions are more
widely adopted, the impact of these efforts are unfortunately limited by the efficiency, motivation
and integrity of their active contributors. Similarly, the success of persistent digital identifier efforts
is conditioned to a large and ubiquitous adoption by both researchers and publishers.
For these reasons, fully automated machine learning-based methods have been proposed during the past decade to
provide immediate, less costly, and satisfactory solutions to author
disambiguation. In this work, our goal is to explore and demonstrate how both
approaches can coexist and benefit from each other.
In particular, we study how labeled data obtained through manual curation (either centralized or
crowdsourced) can be exploited (i) to learn an accurate classifier for
identifying coreferring authors, and (ii) to guide the clustering of scientific
publications by distinct authors in a semi-supervised way.
Our analysis of parameters and features of this large dataset reveal that the general pipeline
commonly used in existing solutions is an effective approach for author disambiguation.
Additionally, we propose alternative strategies for blocking based on the
phonetization of author names to increase recall.

@@

Semi-supervised approaches also allow for the tuning of the
clustering algorithm when the latter is applied to a mixed set of labeled
and unlabeled publications, e.g., by maximizing some clustering performance
metric on the known clusters \cite{levin2012citation}.

@@

Formally, let us assume a set of publications ${\cal P} = \{ p_0, ...,
p_{N-1}\}$ along with the set of unique individuals ${\cal A} = \{ a_0, ...,
a_{M-1}\}$ having together authored all publications in ${\cal P}$.  Let us
define a signature $s \in p$ from a publication as a unique piece of
information identifying one of the authors of $p$ (e.g., the author name, his
affiliation, along with any other metadata that can be derived from $p$, as illustrated in Figure~\ref{fig:signature}). Let us
denote by ${\cal S} = \{ s | s \in p, p \in {\cal P} \}$ the set of all
signatures that can be extracted from all publications in ${\cal P}$.

@@

The simplest and most common strategy for blocking, referred to hereon in as \textit{Surname and First Initial (SFI)},
groups signatures together if they share the same surname(s) and the same first
given name initial (e.g., \emph{SFI}("Doe, John") $==$ "Doe, J").
Despite satisfactory performance, there are several cases where this simple strategy fails to cluster related pairs of signatures together, including:

\begin{enumerate}
  \item There are different
  ways of writing an author name, or signatures contain a typo
  (e.g., "Mueller, R." and "Muller, R.", "Tchaikovsky, P." and "Czajkowski, P.").

  \item An author has multiple surnames and some signatures place the first part of the surname within the given names (e.g., "Martinez Torres, A." and "Torres, A. Martinez").

  \item An author has multiple surnames and, on some signatures, only the first surname is
  present (e.g., "Smith-Jones, A." and "Smith, A.")

  \item An author has multiple given names and they are not always all recorded (e.g.,
  "Smith, Jack" and "Smith, A. J.")

  \item An authors surname changed (e.g., due to marriage).
\end{enumerate}

To account for these issues we propose instead to block signatures based on the
phonetic representation of the normalized surname.
Normalization involves stripping accents (e.g., "Jabłoński, Ł" $\rightarrow$ "Jablonski, L") and name
affixes that inconsistently appear in signatures (e.g., "van der Waals, J. D."
$\rightarrow$ "Waals, J. D."), while phonetization is based either on the
Double Metaphone \cite{doublemetaphone}, the NYSIIS \cite{nysiis} or the
Soundex \cite{Soundex} phonetic algorithms for mapping author names to their pronunciations.
Together, these processing steps allow for grouping of most name variants of the same
person in the same block with a small increase in the overall computational complexity, thereby solving case 1.

@@

\textit{Input features.} In most cases, supervised learning algorithms assume
the input space ${\cal X}$ to be numeric (e.g., $\mathbb{R}^p$), making them
not directly applicable to structured input spaces such as ${\cal S} \times
{\cal S}$.
Following previous works, pairs of signatures $(s_1, s_2)$ are first transformed to vectors $v \in \mathbb{R}^p$
by building so-called similarity profiles \cite{treeratpituk2009disambiguating} on which supervised learning is carried out.
In this work, we design and evaluate fifteen standard input
features based on the comparison of signature fields, as reported in the first
half of Table~\ref{table:features}.
As an illustrative example, the \textit{Full name} feature corresponds to the similarity between the (full)
author name fields of the two signatures, as measured using as combination
operator the cosine similarity between their respective $(n,m)$-\emph{TF-IDF} vector
representations\footnote{$(n,m)$ denotes that the \emph{TF-IDF} vectors are computed
from character $n$, $n+1$, ..., $m$-grams.
When not specified, \emph{TF-IDF} vectors are otherwise computed from words.}.
Similarly, the \textit{Year difference} feature measures the absolute difference between the publication date of the
articles to which the two signatures respectively belong.

Author names from different cultures, origins or ethnic groups are likely to be
disambiguated using different strategies (e.g., pairs of signatures with French
author names versus pairs of signatures with Chinese author names) \cite{treeratpituk2012name, chin2014effective}.
To support our disambiguation algorithm, we added seven features to our feature set, with each
evaluating the degree of belonging of both signatures to an ethnic group,
as reported in the second half of Table~\ref{table:features}.

@@

The most straightforward approach for building a training set on which to learn
a linkage function is to sample an equal number of positive and negative pairs,
as suggested above.
By observing that the linkage function $d$ will eventually
be used only on pairs of signatures from the same block $S_b$, a further
refinement for building a training set is to restrict positive and negative
pairs $(s_1, s_2)$ to only those for which $s_1$ and $s_2$ belong to the same
block. In doing so, the trained classifier is forced to learn intra-block
discriminative patterns rather than inter-block differences.
Furthermore, as noted in \cite{lange2011frequency}, most signature pairs are non-ambiguous:
if both signatures share the same author names, then
they correspond to the same individual, otherwise they do not.
Rather than sampling pairs uniformly at random, we propose to oversample
difficult cases when building the training set (e.g., pairs of signatures with
different author names corresponding to same individual, and pairs of
signatures with identical author names but corresponding to distinct
individuals) in order to improve the overall accuracy of the linkage function.

@@

\textit{Blocking.} The good precision of the baseline ($0.9901$), but its
lower recall ($0.9760$) suggest that the blocking strategy might be the
limiting factor to further overall improvements.
As shown in Table~\ref{table:blocking}, the maximum recall (e.g., if within a block, all signatures were clustered optimally) for SFI is $0.9828$.
At the price of fewer and therefore slightly larger blocks (as reported in the right column of Table~\ref{table:blocking}), the
proposed phonetic-based blocking strategies show better maximum recall (all
around $0.9905$), thereby pushing further the upper bound on the maximum
performance of author disambiguation.
Let us remind however that the reported maximum recalls for the blocking strategies using phonetization are
also raised due to the better handling of multiple surnames, as described in Section \ref{methods:blocking}.

@@

Clustering together all signatures from the same block is the least effective
strategy ($0.9409$), but yields anyhow surprisingly decent accuracy, given the
fact it requires almost no computation (e.g., both learning a linkage function
and running agglomerative clustering can be skipped -- only the blocking
function is needed to group signatures). In particular, this result reveals
that author names are not ambiguous in most cases\footnote{This holds for the
data we extracted, but may in the future, with the rise of non-Western
researchers, be an underestimate of the ambiguous cases.} and that only a small
fraction of them requires advanced disambiguation procedures. On the other
hand, both global and block cut thresholding strategies give very good results,
with a slight advantage for block cuts ($0.9814$ versus $0.9830$), as expected.
In case ${\cal S}^\prime_b$ is empty (e.g., because it corresponds to a young
researcher at the beginning of his career), this therefore suggests that either using
a cut-off threshold learned globally from the known data or using SFI would in general give
satisfactory results, only marginally worse than if claimed signatures had been
known. Trying out 3 clustering methods, the results show that \textit {average linkage} was the best one.

\textit{Combined best settings.} When all best settings are combined (e.g.,
Blocking = NYSIIS, Name normalization, Classifier = Random Forests, Training
pairs = blocked and balanced, Clustering = Average linkage, Block cuts),
performance reaches $0.9862$, e.g., the best of all reported results. In particular,
this combination exhibits both the high recall of phonetic blocking based on the NYSIIS algorithm
and the high precision of Random Forests.

@@

Overall, these results all encourage further improvements and research. For
blocking, one of the open challenges is to manage signatures with inconsistent
surnames or inconsistent first given names (cases 4 and 5, as described in
Section~\ref{methods:blocking}) while maintaining blocks to a tractable size.
As phonetic algorithms are not yet perfect, another direction  for further work is the design of better
phonetization functions, tailored for author disambiguation. For the linkage function,
the good results of the proposed features pave the way for further research  in
ethnicity-sensitive author disambiguation. The automatic fitting of the
pipeline to cultures and ethnic groups for which standard author disambiguation
is known to be less efficient (e.g., Chinese authors with many homonyms)  indeed
constitutes a direction of research with great potential benefits for the
concerned scientific communities.

>>>>

\usepackage{xspace}
\newcommand*{\eg}{e.g.\@\xspace}
\newcommand*{\ie}{i.e.\@\xspace}

@@

\section{Introduction}
In academic digital libraries, author name disambiguation is the problem of
grouping together publications written by the same person.
Author name disambiguation is often a difficult problem because an author may use different
spellings or name variants across their career (synonymy) and/or distinct authors may
share the same name (polysemy).
Most notably, author disambiguation is often more troublesome for researchers
from non-Western cultures, where personal names may be traditionally less diverse (leading to
homonym issues) or for which transliteration to Latin characters may not be unique (leading to
synonym issues).
With the fast growth of the scientific literature, author disambiguation has become a pressing
issue since the accuracy of information managed at the level of individuals directly affects:
the relevance search of results (\eg, when querying for all publications written by a given author);
the reliability of bibliometrics and author rankings (\eg, citation counts or other impact
metrics, as studied in \cite{strotmann2012author}); and/or the relevance of scientific network analysis
\cite{newman2001structure}.

Efforts and solutions to author disambiguation have been proposed from various
communities \cite{liu2014author}. On the one hand, libraries have maintained
authorship control through manual curation, either in a centralized way by
hiring professional collaborators or through developing services that invite authors to register
their publications themselves (\eg, Google Scholar or Inspire-HEP).
Recent efforts to create persistent digital identifiers assigned to researchers (\eg, ORCID or ResearcherID),
with the objective to embed these identifiers in the submission workflow of publishers
or repositories (\eg, Elsevier, arXiv or Inspire-HEP), would univocally solve any disambiguation issue.
With the large cost of centralized manual authorship control, or until crowdsourced solutions are more
widely adopted, the impact of these efforts are unfortunately limited by the efficiency, motivation
and integrity of their active contributors. Similarly, the success of persistent digital identifier efforts
is conditioned to a large and ubiquitous adoption by both researchers and publishers.
For these reasons, fully automated machine learning-based methods have been proposed during the past decade to
provide immediate, less costly, and satisfactory solutions to author
disambiguation. In this work, our goal is to explore and demonstrate how both
approaches can coexist and benefit from each other.
In particular, we study how labeled data obtained through manual curation (either centralized or
crowdsourced) can be exploited (i) to learn an accurate classifier for
identifying coreferring authors, and (ii) to guide the clustering of scientific
publications by distinct authors in a semi-supervised way.
Our analysis of parameters and features of this large dataset reveal that the general pipeline
commonly used in existing solutions is an effective approach for author disambiguation.
Additionally, we propose alternative strategies for blocking based on the
phonetization of author names to increase recall.

@@

Semi-supervised approaches also allow for the tuning of the
clustering algorithm when the latter is applied to a mixed set of labeled
and unlabeled publications, \eg, by maximizing some clustering performance
metric on the known clusters \cite{levin2012citation}.

@@

Formally, let us assume a set of publications ${\cal P} = \{ p_0, ...,
p_{N-1}\}$ along with the set of unique individuals ${\cal A} = \{ a_0, ...,
a_{M-1}\}$ having together authored all publications in ${\cal P}$.  Let us
define a signature $s \in p$ from a publication as a unique piece of
information identifying one of the authors of $p$ (\eg, the author name, his
affiliation, along with any other metadata that can be derived from $p$, as illustrated in Figure~\ref{fig:signature}). Let us
denote by ${\cal S} = \{ s | s \in p, p \in {\cal P} \}$ the set of all
signatures that can be extracted from all publications in ${\cal P}$.

@@

The simplest and most common strategy for blocking, referred to hereon in as \textit{Surname and First Initial (SFI)},
groups signatures together if they share the same surname(s) and the same first
given name initial (\eg, \emph{SFI}("Doe, John") $==$ "Doe, J").
Despite satisfactory performance, there are several cases where this simple strategy fails to cluster related pairs of signatures together, including:

\begin{enumerate}
  \item There are different
  ways of writing an author name, or signatures contain a typo
  (\eg, "Mueller, R." and "Muller, R.", "Tchaikovsky, P." and "Czajkowski, P.").

  \item An author has multiple surnames and some signatures place the first part of the surname within the given names (\eg, "Martinez Torres, A." and "Torres, A. Martinez").

  \item An author has multiple surnames and, on some signatures, only the first surname is
  present (\eg, "Smith-Jones, A." and "Smith, A.")

  \item An author has multiple given names and they are not always all recorded (\eg,
  "Smith, Jack" and "Smith, A. J.")

  \item An authors surname changed (\eg, due to marriage).
\end{enumerate}

To account for these issues we propose instead to block signatures based on the
phonetic representation of the normalized surname.
Normalization involves stripping accents (\eg, "Jabłoński, Ł" $\rightarrow$ "Jablonski, L") and name
affixes that inconsistently appear in signatures (\eg, "van der Waals, J. D."
$\rightarrow$ "Waals, J. D."), while phonetization is based either on the
Double Metaphone \cite{doublemetaphone}, the NYSIIS \cite{nysiis} or the
Soundex \cite{Soundex} phonetic algorithms for mapping author names to their pronunciations.
Together, these processing steps allow for grouping of most name variants of the same
person in the same block with a small increase in the overall computational complexity, thereby solving case 1.

@@

\textit{Input features.} In most cases, supervised learning algorithms assume
the input space ${\cal X}$ to be numeric (\eg, $\mathbb{R}^p$), making them
not directly applicable to structured input spaces such as ${\cal S} \times
{\cal S}$.
Following previous works, pairs of signatures $(s_1, s_2)$ are first transformed to vectors $v \in \mathbb{R}^p$
by building so-called similarity profiles \cite{treeratpituk2009disambiguating} on which supervised learning is carried out.
In this work, we design and evaluate fifteen standard input
features based on the comparison of signature fields, as reported in the first
half of Table~\ref{table:features}.
As an illustrative example, the \textit{Full name} feature corresponds to the similarity between the (full)
author name fields of the two signatures, as measured using as combination
operator the cosine similarity between their respective $(n,m)$-\emph{TF-IDF} vector
representations\footnote{$(n,m)$ denotes that the \emph{TF-IDF} vectors are computed
from character $n$, $n+1$, ..., $m$-grams.
When not specified, \emph{TF-IDF} vectors are otherwise computed from words.}.
Similarly, the \textit{Year difference} feature measures the absolute difference between the publication date of the
articles to which the two signatures respectively belong.

Author names from different cultures, origins or ethnic groups are likely to be
disambiguated using different strategies (\eg, pairs of signatures with French
author names versus pairs of signatures with Chinese author names) \cite{treeratpituk2012name, chin2014effective}.
To support our disambiguation algorithm, we added seven features to our feature set, with each
evaluating the degree of belonging of both signatures to an ethnic group,
as reported in the second half of Table~\ref{table:features}.

@@

The most straightforward approach for building a training set on which to learn
a linkage function is to sample an equal number of positive and negative pairs,
as suggested above.
By observing that the linkage function $d$ will eventually
be used only on pairs of signatures from the same block $S_b$, a further
refinement for building a training set is to restrict positive and negative
pairs $(s_1, s_2)$ to only those for which $s_1$ and $s_2$ belong to the same
block. In doing so, the trained classifier is forced to learn intra-block
discriminative patterns rather than inter-block differences.
Furthermore, as noted in \cite{lange2011frequency}, most signature pairs are non-ambiguous:
if both signatures share the same author names, then
they correspond to the same individual, otherwise they do not.
Rather than sampling pairs uniformly at random, we propose to oversample
difficult cases when building the training set (\ie, pairs of signatures with
different author names corresponding to same individual, and pairs of
signatures with identical author names but corresponding to distinct
individuals) in order to improve the overall accuracy of the linkage function.

@@

\textit{Blocking.} The good precision of the baseline ($0.9901$), but its
lower recall ($0.9760$) suggest that the blocking strategy might be the
limiting factor to further overall improvements.
As shown in Table~\ref{table:blocking}, the maximum recall (\ie, if within a block, all signatures were clustered optimally) for SFI is $0.9828$.
At the price of fewer and therefore slightly larger blocks (as reported in the right column of Table~\ref{table:blocking}), the
proposed phonetic-based blocking strategies show better maximum recall (all
around $0.9905$), thereby pushing further the upper bound on the maximum
performance of author disambiguation.
Let us remind however that the reported maximum recalls for the blocking strategies using phonetization are
also raised due to the better handling of multiple surnames, as described in Section \ref{methods:blocking}.

@@

Clustering together all signatures from the same block is the least effective
strategy ($0.9409$), but yields anyhow surprisingly decent accuracy, given the
fact it requires almost no computation (\ie, both learning a linkage function
and running agglomerative clustering can be skipped -- only the blocking
function is needed to group signatures). In particular, this result reveals
that author names are not ambiguous in most cases\footnote{This holds for the
data we extracted, but may in the future, with the rise of non-Western
researchers, be an underestimate of the ambiguous cases.} and that only a small
fraction of them requires advanced disambiguation procedures. On the other
hand, both global and block cut thresholding strategies give very good results,
with a slight advantage for block cuts ($0.9814$ versus $0.9830$), as expected.
In case ${\cal S}^\prime_b$ is empty (\eg, because it corresponds to a young
researcher at the beginning of his career), this therefore suggests that either using
a cut-off threshold learned globally from the known data or using SFI would in general give
satisfactory results, only marginally worse than if claimed signatures had been
known. Trying out 3 clustering methods, the results show that \textit {average linkage} was the best one.

\textit{Combined best settings.} When all best settings are combined (\ie, Blocking = NYSIIS, Name normalization, Classifier = Random Forests, Training
pairs = blocked and balanced, Clustering = Average linkage, Block cuts),
performance reaches $0.9862$, \ie, the best of all reported results. In particular,
this combination exhibits both the high recall of phonetic blocking based on the NYSIIS algorithm
and the high precision of Random Forests.

@@

Overall, these results all encourage further improvements and research. For
blocking, one of the open challenges is to manage signatures with inconsistent
surnames or inconsistent first given names (cases 4 and 5, as described in
Section~\ref{methods:blocking}) while maintaining blocks to a tractable size.
As phonetic algorithms are not yet perfect, another direction  for further work is the design of better
phonetization functions, tailored for author disambiguation. For the linkage function,
the good results of the proposed features pave the way for further research  in
ethnicity-sensitive author disambiguation. The automatic fitting of the
pipeline to cultures and ethnic groups for which standard author disambiguation
is known to be less efficient (\eg, Chinese authors with many homonyms)  indeed
constitutes a direction of research with great potential benefits for the
concerned scientific communities.
